---
title: "CSIAPPS REST API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CSIAPPS REST API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(csiapps)
```

### Preamble

The following environment variables must be set to make requests to the `CSIAPPS` REST API:

-   `CSIAPPS_ACCESS_TOKEN`: The access token for authenticating API requests. This token should have the necessary permissions to access the desired endpoints.

### Overview

`csiapps` includes a generic `make_request()` function to make requests to the `CSIAPPS` REST API.
Please refer to the [CSIAPPS Swagger documentation](https://apps.csiontario.ca/api/swagger/){target="_blank"}
for full details on available endpoints, parameters, and response formats.

Below are some common use cases for interacting with the API endpoints using `csiapps`.

> **Note:** By default, `csiapps` runs in **sandbox mode**, so the `make_request()`
> calls below are routed to a local, in-memory warehouse rather than the production
> API. This lets you develop and test safely without touching production data. See
> [Sandbox Mode](#sandbox-mode) below for how it works and how it is turned off for
> deployment.

### Authorization Status

The most basic endpoint is the `/api/csiauth/me/` endpoint, which can be used to check the
status of the user making the request. This endpoint requires authentication, so if the request is successful, it indicates that your access token is valid and has the necessary permissions to access the REST API.
You may need to specify your institute before authenticating.

```{r}

set_institute("csiontario") # or "csipacific" (the default)

result <- make_request(
  endpoint = "api/csiauth/me/"
)
```


### Data Warehouse Ingestion

To ingest data records into the warehouse, we make a POST request to the `/api/warehouse/ingestion/primary/` endpoint.
Users supply the `source` (uuid) of the target *data source*,
a list of `records` to be ingested, and (optionally) the `subject_field` by which each record can be uniquely assigned to a user. The records must comply with the schema defined for the target data source.


#### JSON Schema Definition

Consider the following [JSON Schema](https://json-schema.org/learn){target="_blank"}:

``` json
{
  "title": "A registration form",
  "description": "A simple form example.",
  "type": "object",
  "required": [
    "id",
    "firstName",
    "lastName"
  ],
  "properties": {
    "id": {
      "type": "string",
      "title": "ID"
    },
    "firstName": {
      "type": "string",
      "title": "First name",
      "default": "Chuck"
    },
    "lastName": {
      "type": "string",
      "title": "Last name",
      "default": "Norris"
    },
    "age": {
      "type": "integer",
      "title": "Age"
    },
    "telephone": {
      "type": "string",
      "title": "Telephone",
      "minLength": 10
    }
  }
}
```
Here, `id` is the required field that serves as the unique user assignment for each record.

To create records that comply with this schema, we create the following records:

```{r}
records <- list(
  list(id = "xxxx", firstName = "John", lastName = "Doe", age = 30, telephone = "1234567890"),
  list(id = "yyyy", firstName = "Jane", lastName = "Smith", age = 25, telephone = "0987654321")
)
```

To retrieve the JSON schema definition from the data warehouse, we can make a request to the `/api/warehouse/data-sources/{uuid}/` endpoint, where `uuid` is the uuid of the target data source. 
The schema definition is located in the `head_primary_definition$schema` field of the response.

```{r}
data_source <- make_request(
  endpoint = paste0("api/warehouse/data-sources/", Sys.getenv("SOURCE_UUID"))
)

schema <- data_source$head_primary_definition$schema
```

#### Data Record Validation

We can validate data records using the `jsonlite` and `jsonvalidate` packages. 
Be sure to set `auto_unbox = TRUE` when converting the JSON objects to strings.

```{r}
json_schema <- jsonvalidate::json_schema$new(jsonlite::toJSON(schema, auto_unbox = T))

validate_record <- function(record) {
  json_schema$validate(jsonlite::toJSON(record, auto_unbox = T))
}

validation_results <- sapply(records, validate_record)

stopifnot(all(validation_results))
```

Finally, we can make the API request to ingest these records into the warehouse:

```{r}
result <- make_request(
  endpoint = "api/warehouse/ingestion/primary/",
  method = "POST",
  body = list(
    source = Sys.getenv("SOURCE_UUID"),
    records = records,
    subject_field = "id"
  )
)
```

### Data Record Retrieval 

To retrieve data records from the warehouse, we make an API request to the `/api/warehouse/data-records/` endpoint.
The only required parameter is the `source_uuid` of the requested table, but there are several optional parameters for additional filtering. Note that we can set the `paginate = TRUE` to retrieve all records matching the query parameters.

```{r list-records}
records <- make_request(
    endpoint = "api/warehouse/data-records",
    query = list(source_uuid = Sys.getenv("SOURCE_UUID")),
    paginate = TRUE
    )
```

### Sandbox Mode

`csiapps` includes a local sandbox that emulates the warehouse endpoints above, so
ingestion pipelines can be developed and tested with no network access, no
authentication, and no risk of writing test data to the production warehouse.
Sandbox requests are routed to an in-memory warehouse that validates records
against a registered JSON schema (using the `jsonvalidate` package) and stores
accepted payloads locally.

**Sandbox mode is enabled by default.** Every one of the `make_request()` calls
shown earlier in this vignette runs against the local sandbox until sandbox mode
is explicitly turned off. Every sandboxed request also prints a
`csiapps sandbox:` message, so it is always visible that no real API call was made.

To turn sandbox mode off — for example, when deploying a finished app to
production — set the R option to `FALSE`, or the `CSIAPPS_ENV` environment
variable to `"production"`:

```{r}
options(csiapps.sandbox = FALSE) # route requests to the production warehouse
```

The `sandbox` argument of `make_request()` can also override the global setting
for an individual call, e.g. `make_request(..., sandbox = FALSE)`.

#### Best Practices

Follow these two practices so that apps transition cleanly from sandbox to
production:

1.  **Do not set the `csiapps.sandbox` option (or the `sandbox` argument) in your
    app code.** Rely on the default. Because sandbox mode is on by default, your
    app runs safely in the sandbox during development without any configuration.
    Hard-coding `options(csiapps.sandbox = TRUE)` in committed code would override
    the deployer's attempt to turn sandbox off, causing the deployed app to
    silently serve sandbox data.

2.  **Turn sandbox off only at deployment, in a location the deployer controls.**
    Sandbox mode should be disabled when the finished app is deployed — typically
    by setting `options(csiapps.sandbox = FALSE)` in the app's `global.R` (or an
    `Rprofile.site`) at deployment time, or by setting `CSIAPPS_ENV=production` in
    the production environment. Keeping this out of your app code means the same
    code runs in the sandbox during development and against production once
    deployed, with no edits in between.

#### Registering a Data Source

The sandbox needs a schema to validate against. Register one under any
identifying string — a real production uuid is not required:

```{r}
register_sandbox_schema("dummy-1234", "path/to/schema.json")
```

The schema may be an R list, a JSON string, or a path to a JSON file.

#### The Sandbox Workflow

From here, the workflow is *identical* to the production examples above — the
same `make_request()` calls, with the dummy uuid in place of `Sys.getenv("SOURCE_UUID")`:

```{r}
# Retrieve the registered schema
data_source <- make_request(
  endpoint = paste0("api/warehouse/data-sources/", "dummy-1234")
)
schema <- data_source$head_primary_definition$schema

# Ingest records (validated against the schema before acceptance)
result <- make_request(
  endpoint = "api/warehouse/ingestion/primary/",
  method = "POST",
  body = list(
    source = "dummy-1234",
    records = records,
    subject_field = "id"
  )
)

# Retrieve ingested records
records <- make_request(
  endpoint = "api/warehouse/data-records",
  query = list(source_uuid = "dummy-1234"),
  paginate = TRUE
)
```

Records that fail schema validation are rejected with an error describing
exactly which records failed and why. When code is ready for production,
disable sandbox mode and swap in the real source uuid — no other changes are
needed.

#### Inspecting and Resetting the Sandbox

Accepted payloads are written as pretty-printed JSON files, one folder per data
source, so the exact JSON that would have been sent to the API can be inspected:

```{r}
browse_sandbox("dummy-1234") # opens the payload folder in the file explorer
```

To reset state between tests, clear a single source or the entire sandbox:

```{r}
clear_sandbox("dummy-1234") # clear one source
clear_sandbox()             # clear everything
```

#### Limitations

The sandbox faithfully simulates the *schema contract*, not the warehouse.
Anything that depends on server-side state will differ from production:

-   **Validation parity is approximate.** The sandbox validates records against
    the JSON Schema only. The real server additionally enforces things the
    schema cannot express: resolution of `subject_field` values against
    registered profiles, duplicate handling against existing data, and token
    permissions. Passing sandbox validation is a *necessary but not sufficient*
    condition for production acceptance.
-   **`subject` is always `NULL` in retrieved records.** The sandbox has no
    profile registry, so it cannot emulate the server's record-to-profile
    linkage. Code that displays or filters on subject fields will see
    placeholder values in sandbox mode.

See `?"csiapps-sandbox"` for full details.
