---
title: "CSIAPPS REST API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CSIAPPS REST API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(csiapps)
```

### Preamble

The following environment variables must be set to make requests to the `CSIAPPS` REST API:

-   `CSIAPPS_ACCESS_TOKEN`: The access token for authenticating API requests. This token should have the necessary permissions to access the desired endpoints.

### Overview

`csiapps` includes a generic `make_request()` function to make requests to the `CSIAPPS` REST API.
Please refer to the [CSIAPPS Swagger documentation](https://apps.csiontario.ca/api/swagger/){target="_blank"}
for full details on available endpoints, parameters, and response formats.

Below are some common use cases for interacting with the API endpoints using `csiapps`.

> **Note:** By default, `csiapps` runs in **sandbox mode**, so the `make_request()`
> calls below are routed to a local, in-memory warehouse rather than the production
> API. This lets you develop and test safely without touching production data. This
> article documents the **production** REST API semantics; for the full local
> workflow — registering orgs and athletes, schemas, ingestion, and retrieval
> against the emulated warehouse — see the dedicated [Sandbox Mode](sandbox.html)
> article.

### Authorization Status

The most basic endpoint is the `/api/csiauth/me/` endpoint, which can be used to check the
status of the user making the request. This endpoint requires authentication, so if the request is successful, it indicates that your access token is valid and has the necessary permissions to access the REST API.
You may need to specify your institute before authenticating.

```{r}

set_institute("csiontario") # or "csipacific" (the default)

result <- make_request(
  endpoint = "api/csiauth/me/"
)
```


### Registration API

The registration API provides access to profiles and organisations via
`fetch_org_options()`, `fetch_profiles()`, and `fetch_profile()`. The examples
below pass `sandbox = FALSE` to contact the **real** registration API. In
sandbox mode (the default) these same helpers read from a local dummy registry —
see the [Sandbox Mode](sandbox.html) article for that workflow.

#### Organisations

`fetch_org_options()` returns all organisations your token can see, formatted
as `label`/`value` pairs ready for a Shiny `selectInput()`:

```{r}
set_institute("csiontario")

orgs <- fetch_org_options(sandbox = FALSE)
# list(list(label = "CSI Ontario", value = 1L), ...)
```

#### Profiles

`fetch_profiles()` returns all profiles accessible to your token, with
automatic pagination. Filter by organisation using `sport_org_id`:

```{r}
profiles <- fetch_profiles(sandbox = FALSE)

# Filter to a specific organisation
profiles <- fetch_profiles(
  filters = list(sport_org_id = 42L),
  sandbox = FALSE
)
```

A common pattern across dashboards is converting the profile list into a data
frame for display or matching:

```{r}
profile_df <- do.call(rbind, lapply(profiles, function(p) {
  data.frame(
    id         = p$id,
    first_name = p$person$first_name %||% NA_character_,
    last_name  = p$person$last_name  %||% NA_character_,
    email      = p$person$email      %||% NA_character_
  )
}))
```

`fetch_profile()` retrieves a single profile by ID:

```{r}
profile <- fetch_profile(profile_id = 123L, sandbox = FALSE)
profile$person$first_name
```

### Data Warehouse Ingestion

To ingest data records into the warehouse, we make a POST request to the `/api/warehouse/ingestion/primary/` endpoint.
Users supply the `source` (uuid) of the target *data source*,
a list of `records` to be ingested, and (optionally) the `subject_field` by which each record can be uniquely assigned to a user. The records must comply with the schema defined for the target data source.

> **Note:** The `source` uuid is provisioned by the CSIAPPS team when they set up
> the remote data warehouse for your app, and supplied through the `SOURCE_UUID`
> environment variable — the app developer does not choose or hard-code it. Read
> it as `Sys.getenv("SOURCE_UUID")` (as below) so the same code runs in
> development and production. See the [Sandbox Mode](sandbox.html) article for how
> this works during local development.


#### JSON Schema Definition

Consider the following [JSON Schema](https://json-schema.org/learn){target="_blank"}:

``` json
{
  "title": "A registration form",
  "description": "A simple form example.",
  "type": "object",
  "required": [
    "id",
    "firstName",
    "lastName"
  ],
  "properties": {
    "id": {
      "type": "string",
      "title": "ID"
    },
    "firstName": {
      "type": "string",
      "title": "First name",
      "default": "Chuck"
    },
    "lastName": {
      "type": "string",
      "title": "Last name",
      "default": "Norris"
    },
    "age": {
      "type": "integer",
      "title": "Age"
    },
    "telephone": {
      "type": "string",
      "title": "Telephone",
      "minLength": 10
    }
  }
}
```
Here, `id` is the required field that serves as the unique user assignment for each record.

To create records that comply with this schema, we create the following records:

```{r}
records <- list(
  list(id = "xxxx", firstName = "John", lastName = "Doe", age = 30, telephone = "1234567890"),
  list(id = "yyyy", firstName = "Jane", lastName = "Smith", age = 25, telephone = "0987654321")
)
```

To retrieve the JSON schema definition from the data warehouse, we can make a request to the `/api/warehouse/data-sources/{uuid}/` endpoint, where `uuid` is the uuid of the target data source. 
The schema definition is located in the `head_primary_definition$schema` field of the response.

```{r}
data_source <- make_request(
  endpoint = paste0("api/warehouse/data-sources/", Sys.getenv("SOURCE_UUID"))
)

schema <- data_source$head_primary_definition$schema
```

#### Data Record Validation

We can validate data records using the `jsonlite` and `jsonvalidate` packages. 
Be sure to set `auto_unbox = TRUE` when converting the JSON objects to strings.

```{r}
json_schema <- jsonvalidate::json_schema$new(jsonlite::toJSON(schema, auto_unbox = T))

validate_record <- function(record) {
  json_schema$validate(jsonlite::toJSON(record, auto_unbox = T))
}

validation_results <- sapply(records, validate_record)

stopifnot(all(validation_results))
```

Finally, we can make the API request to ingest these records into the warehouse:

```{r}
result <- make_request(
  endpoint = "api/warehouse/ingestion/primary/",
  method = "POST",
  body = list(
    source = Sys.getenv("SOURCE_UUID"),
    records = records,
    subject_field = "id"
  )
)
```

### Data Record Retrieval 

To retrieve data records from the warehouse, we make an API request to the `/api/warehouse/data-records/` endpoint.
The only required parameter is the `source_uuid` of the requested table, but there are several optional parameters for additional filtering. Note that we can set the `paginate = TRUE` to retrieve all records matching the query parameters.

```{r list-records}
records <- make_request(
    endpoint = "api/warehouse/data-records",
    query = list(source_uuid = Sys.getenv("SOURCE_UUID")),
    paginate = TRUE
    )
```

### Sandbox Mode

Every `make_request()` call shown above runs against a **local sandbox** by
default, so ingestion and retrieval pipelines can be developed and tested with no
network access, no authentication, and no risk of writing test data to the
production warehouse. Sandbox mode is enabled by default and is turned off only
at deployment.

The full sandbox workflow — enabling/disabling it, registering sport orgs and
athletes, registering schemas, validating, ingesting, and retrieving records
against the emulated warehouse, and inspecting or resetting sandbox state — is
documented in its own article:

**→ [Sandbox Mode](sandbox.html)**

See also `?"csiapps-sandbox"` for the reference-level overview.
