---
title: "Developing Shiny Web Applications"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Developing Shiny Web Applications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

### Preamble

By default, `csiapps` runs in **sandbox mode**, so a wrapped app can be developed
and run locally *without* any of the OAuth client credentials below: the login
redirect is simulated (see [Sandbox mode](#sandbox-mode) at the end of this
article). The following environment variables are only required for a real
deployment, once sandbox mode is turned off:

-   `CSIAPPS_CLIENT_ID`: The client ID for the application registered in CSIAPPS.
-   `CSIAPPS_CLIENT_SECRET`: The client secret for the application registered in CSIAPPS.
-   `CSIAPPS_REDIRECT_URL`: The URL to which the application will redirect after authentication.
-   `CSIAPPS_SCOPE`: (optional) The scope of the authentication request.

## Example
Suppose that you have the following shiny web application:

```{r}
library(shiny)

df = faithful[, 2]

ui <- fluidPage(

    titlePanel("Old Faithful Geyser Data"),

    sidebarLayout(
        sidebarPanel(
            sliderInput("bins",
                        "Number of bins:",
                        min = 1,
                        max = 50,
                        value = 30)
        ),

        mainPanel(
           plotOutput("distPlot")
        )
    )
)

server <- function(input, output) {

    output$distPlot <- renderPlot({
        bins <- seq(min(df), max(df), length.out = input$bins + 1)

        hist(df, breaks = bins, col = 'darkgray', border = 'white',
             xlab = 'Waiting time to next eruption (in mins)',
             main = 'Histogram of waiting times')
    })
}

shinyApp(ui = ui, server = server)
```


To migrate this app within the `CSIAPPS` ecosystem, we can leverage several functions provided by `csiapps`.

### 1. `set_institute()`

We specify which institute internal API calls should be made for (such as authentication redirects) by the application using `set_institute()`.

```{r set-institute}
# institute can be set to "csiontario", "csipacific", or "csiatlantic"
csiapps::set_institute("csiontario")
```

### 2. `check_secrets()`

We then run `check_secrets()` to ensure that all environment variables required have been set. The `verbose` argument, which is `FALSE` by default, can be set to `TRUE` to print out the values of the environment variables that are being checked. `check_secrets()` will throw an error if any of the required environment variables are not present, making it useful for debugging.

```{r check-secrets}
csiapps::check_secrets(verbose = FALSE)
```

### 3. `global_wrapper()`

For code defined outside but used within the `server` function, we use
`global_wrapper()` to ensure that it is accessible by internal helper functions.

```{r global-wrapper}
csiapps::global_wrapper({
  df = faithful[, 2]
})
```

### 4. `ui_wrapper()` and `server_wrapper()`

The `ui` page and `server` function can simply be wrapped by `ui_wrapper()` and `server_wrapper()`, respectively.
These convenience functions include additional 
code to redirect the application to CSIAPPS for user authentication and provide aesthetic formatting.

```{r ui-wrapper}
# original code

ui <- ...

server <- ...

shinyApp(ui = ui, server = server)

# wrapped code

shinyApp(ui = csiapps::ui_wrapper(ui), server = csiapps::server_wrapper(server))
```

### Summary

Therefore, the full code for the app, after migration, would look like this:

```{r}
library(shiny)
library(csiapps)

# CSIAPPS Setup
set_institute("csiontario")
check_secrets()

global_wrapper({
  df = faithful[, 2]
})

# Define UI for application that draws a histogram
ui <- fluidPage(

    titlePanel("Old Faithful Geyser Data"),

    sidebarLayout(
        sidebarPanel(
            sliderInput("bins",
                        "Number of bins:",
                        min = 1,
                        max = 50,
                        value = 30)
        ),

        mainPanel(
           plotOutput("distPlot")
        )
    )
)

# Define server logic required to draw a histogram
server <- function(input, output) {

    output$distPlot <- renderPlot({
        bins <- seq(min(df), max(df), length.out = input$bins + 1)

        # draw the histogram with the specified number of bins
        hist(df, breaks = bins, col = 'darkgray', border = 'white',
             xlab = 'Waiting time to next eruption (in mins)',
             main = 'Histogram of waiting times')
    })
}

# Run the application
shinyApp(ui = ui_wrapper(ui), server = server_wrapper(server))
```

### Sandbox mode

The code above is written for production, but you do **not** need any client
credentials or a running identity provider to develop it. By default `csiapps`
is in **sandbox mode**, and `server_wrapper()` **simulates the login redirect**:
instead of sending the browser to CSIAPPS to authenticate, it seeds the session
from the `CSIAPPS_ACCESS_TOKEN` you already have in your environment.

```{r}
Sys.setenv(CSIAPPS_ACCESS_TOKEN = "your-dev-access-token")
```

With that token set, the wrapped app runs locally and the header loads your
**real** identity (`/me`) from the API, exactly as it would after a production
login. The token is used **only to emulate the login** — everything else (sport
organizations, athletes, warehouse records) is dummy, served from the local
sandbox (see `create_sport_org()`, `create_profile()`, and `vignette("api")`).
`check_secrets()` reports whether a token was found rather than erroring on the
(unneeded) OAuth secrets. If no token is set, the app shell still renders but
shows an unauthenticated notice prompting you to set a read-only
`CSIAPPS_ACCESS_TOKEN`.

> **Note:** Sandbox mode never touches real client data. Warehouse calls made
> through `make_request()` are emulated locally, and registration reads
> (`fetch_org_options()`, `fetch_profiles()`) come from the local dummy
> registry. Only `/me` uses your real token — call `set_institute()` to match
> the institute that issued it, or that lookup is rejected.

### Deploying

Because the app code is identical in both modes, deploying is just a matter of
turning sandbox mode off — no code changes required:

```{r}
options(csiapps.sandbox = FALSE) # or set CSIAPPS_ENV=production
```

Once your app is wrapped and ready to deploy, generate a `manifest.json` from
the app directory. The deployment target (Posit Connect / shinyapps.io) uses it
to reproduce your package environment — including `csiapps` — so it must be
regenerated whenever your dependencies change:

```{r}
rsconnect::writeManifest()
```

Commit the resulting `manifest.json` alongside your app.
