---
title: "Sandbox Mode"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sandbox Mode}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(csiapps)
```

## What sandbox mode is

`csiapps` ships with a **local sandbox** that emulates the CSIAPPS data
warehouse and registration API entirely on your machine — no network access, no
authentication, and no risk of writing test data to production. It lets you run
the full **schema → validate → ingest → retrieve** warehouse workflow, plus the
**register orgs → register athletes → fetch profiles** registration workflow,
using the *same* function calls your production app will use.

**Sandbox mode is enabled by default.** This is deliberate: development work
never reaches the production warehouse by accident. Every request is routed to a
local, in-memory warehouse instead of the REST API, and every sandboxed call
prints a `csiapps sandbox:` message so it is always visible that no real API
call was made. When your app is ready, you disable sandbox mode at deployment
time — with no changes to your app code.

This article walks through sandbox mode from start to finish. If you are looking
for the production REST API semantics, see the
[CSIAPPS REST API](api.html) article; for wrapping a Shiny app, see
[Developing Shiny Web Applications](csiapps.html).

### What is real and what is dummy

Sandbox mode is careful about the one thing it cannot fake safely — your
identity — versus everything else, which is dummy:

| Concern | In sandbox mode |
| --- | --- |
| **Login / `/me` identity** | Emulated using your real `CSIAPPS_ACCESS_TOKEN`. The token is used *only* to load `/me`, so a wrapped app's header shows your real name. |
| **Sport organizations** | Dummy. Created locally with `create_sport_org()`. |
| **Athlete profiles** | Dummy. Created locally with `create_profile()` (names drawn from the `babynames` dataset). |
| **Warehouse schemas & records** | Dummy. Registered and ingested into an in-memory warehouse; nothing leaves your machine. |

No real client data is ever read except your own `/me` identity.

## Enabling and disabling sandbox mode

`is_sandbox_mode()` reports whether sandbox mode is currently on, and it supplies
the default for the `sandbox` argument of every sandbox-aware function.

```{r}
is_sandbox_mode() # TRUE by default
```

The resolution order is:

1. The **`csiapps.sandbox` R option**, if set, always wins.
2. Otherwise, the **`CSIAPPS_ENV`** environment variable — only the exact value
   `"production"` disables the sandbox.
3. Otherwise, sandbox mode is **on**.

```{r}
# Turn sandbox OFF (route to the production warehouse)
options(csiapps.sandbox = FALSE)   # ...or set CSIAPPS_ENV=production

# Turn sandbox back ON / restore the default
options(csiapps.sandbox = NULL)
```

You can also override the global setting for a **single call** using the
`sandbox` argument, which every relevant function accepts. This is handy when you
want one real lookup while otherwise developing in the sandbox:

```{r}
# One real org lookup while the sandbox is otherwise on
orgs <- fetch_org_options(sandbox = FALSE)
```

### Best practices for the sandbox → production transition

Follow these two rules so apps move cleanly from development to deployment:

1.  **Do not set the `csiapps.sandbox` option (or the `sandbox` argument) in your
    app code.** Rely on the default. Because sandbox mode is on by default, your
    app runs safely in the sandbox during development with no configuration.
    Hard-coding `options(csiapps.sandbox = TRUE)` in committed code would
    override the deployer's attempt to turn the sandbox off, causing a deployed
    app to silently serve sandbox data.

2.  **Turn sandbox off only at deployment, in a location the deployer controls** —
    typically `options(csiapps.sandbox = FALSE)` in the app's `global.R` (or an
    `Rprofile.site`), or by setting `CSIAPPS_ENV=production` in the production
    environment. Keeping this out of your app code means the same code runs in
    the sandbox during development and against production once deployed, with no
    edits in between.

## Running a wrapped Shiny app in the sandbox

Sandbox mode also lets a wrapped Shiny app (see [`ui_wrapper()`](../reference/ui_wrapper.html),
[`server_wrapper()`](../reference/server_wrapper.html)) run locally **without**
the OAuth2 redirect. The redirect exists only to obtain an access token, and it
requires client credentials that cannot be safely distributed — so in sandbox
mode `server_wrapper()` **simulates the login** instead: it seeds the session
from the `CSIAPPS_ACCESS_TOKEN` you already have and hands it to the same code
path a production login would use.

```{r}
Sys.setenv(CSIAPPS_ACCESS_TOKEN = "your-dev-access-token")

set_institute("csiontario") # must match the institute that issued the token
check_secrets()             # reports whether a token was found (never errors in sandbox)
```

With a token set, the app's header loads your **real** identity from `/me`,
exactly as after a production login. Because `/me` is the *only* real call, you
must `set_institute()` to match the institute that issued the token, or the
lookup is rejected.

If **no** token is set, the app shell still renders but shows an unauthenticated
notice prompting you to set a read-only `CSIAPPS_ACCESS_TOKEN`. In sandbox mode
`check_secrets()` never errors on the (unneeded) OAuth secrets — it simply
reports whether a token was found.

## The registration workflow

In sandbox mode the registration helpers — `fetch_org_options()`,
`fetch_profiles()`, and `fetch_profile()` — read from a **local dummy registry**
instead of the network. You seed that registry with `create_sport_org()` and
`create_profile()`, and the fetch helpers then return dummy data in the same
shape the real API uses, so dashboards can be built and tested without a live
connection.

### 1. Register sport organizations

`create_sport_org()` registers a dummy sport org. Its `name` becomes the
`sport$name` of every athlete you later create under it. If you omit `id`, an
unused id in `1:999` is generated; otherwise you may pin any positive integer
id up to and including 999.

```{r}
# Auto-generated id
org <- create_sport_org("Rowing Canada")
org$id

# ...or pin a specific id
create_sport_org("Athletics Canada", id = 42L)
```

Ids must be unique within the sandbox; reusing one raises an error. Calling
`create_sport_org()` outside sandbox mode warns and has no effect (dummy orgs are
only read by the sandbox helpers).

### 2. Register athletes under a sport org

`create_profile()` generates `n` random athlete profiles and registers them under
an **existing** sport org. Each profile is production-shaped (a `person`
sub-list, a `sport` link, status fields, and so on), so downstream code sees the
same structure the real API returns.

```{r}
# 5 athletes with unique random names drawn from `babynames`
create_profile(5, org$id)

# ...or supply explicit names (provide both vectors, or neither)
create_profile(
  2, org$id,
  first_names = c("Ada", "Blair"),
  last_names  = c("Nkemelu", "Okafor")
)
```

The org must already exist — `create_profile()` errors if `sport_org_id` is
unknown. Profiles are appended to any already registered, and each athlete's
`sport$id` is the `sport_org_id`, which is the field the `sport_org_id` filter
matches against in `fetch_profiles`.

### 3. Fetch orgs and athlete profiles

Once the registry is seeded, the same fetch helpers your production app uses read
it back. `fetch_org_options()` returns `label`/`value` pairs ready for a Shiny
`selectInput()`:

```{r}
fetch_org_options()   # list(list(label = "Rowing Canada", value = 123L), ...)
```

`fetch_profiles()` returns all registered athletes, optionally filtered by
organisation. **Only the `sport_org_id` filter is honoured in the sandbox**;
other filters are ignored.

```{r}
# All dummy athletes
profiles <- fetch_profiles()

# Athletes for one organisation
profiles <- fetch_profiles(filters = list(sport_org_id = org$id))
```

A common pattern is flattening the profile list into a data frame for display:

```{r}
profile_df <- do.call(rbind, lapply(profiles, function(p) {
  data.frame(
    id         = p$id,
    first_name = p$person$first_name %||% NA_character_,
    last_name  = p$person$last_name  %||% NA_character_,
    email      = p$person$email      %||% NA_character_
  )
}))
```

`fetch_profile()` retrieves a single athlete by id, returning `NULL` if no such
id is registered:

```{r}
profile <- fetch_profile(profile_id = 1L)
profile$person$first_name
```

With nothing registered, `fetch_org_options()` and `fetch_profiles()` return
`list()` and `fetch_profile()` returns `NULL`.

> **Note:** Only the warehouse endpoints are routed through `make_request()`.
> Calling `make_request("api/registration/...")` in the sandbox raises a 501 —
> use the `fetch_*` helpers above, which read the dummy registry instead.

## The data warehouse workflow

The sandbox emulates three warehouse endpoints through the ordinary
`make_request()` interface, validating records against a registered JSON schema
and storing accepted payloads locally:

* `GET  api/warehouse/data-sources/{uuid}` — returns a registered schema.
* `POST api/warehouse/ingestion/primary/` — validates and stores records.
* `GET  api/warehouse/data-records` — returns previously ingested records.

### 1. Register a data source (schema)

> **You never handle the real data source uuid.** In production, a CSIAPPS
> insider provisions the remote warehouse and establishes the source uuid for
> your app, supplying it through the `SOURCE_UUID` environment variable. Your app
> code should therefore reference the source **only** as `Sys.getenv("SOURCE_UUID")`
> — never a hard-coded uuid — so the same code runs unchanged in development and
> in production. During development you set `SOURCE_UUID` yourself to any
> throwaway placeholder; its exact value is irrelevant to the sandbox.

Set a placeholder for development and register a schema under it. The sandbox
accepts any identifying string as the source — a real production uuid is **not**
required. The schema may be an R list, a JSON string, or a path to a JSON file.

```{r}
Sys.setenv(SOURCE_UUID = "dev-placeholder") # dev only; CSIAPPS sets this in production

schema <- '{
  "title": "A registration form",
  "type": "object",
  "required": ["id", "firstName", "lastName"],
  "properties": {
    "id":        {"type": "string",  "title": "ID"},
    "firstName": {"type": "string",  "title": "First name"},
    "lastName":  {"type": "string",  "title": "Last name"},
    "age":       {"type": "integer", "title": "Age"},
    "telephone": {"type": "string",  "title": "Telephone", "minLength": 10}
  }
}'

register_sandbox_schema(Sys.getenv("SOURCE_UUID"), schema)
```

Here `id` is the field that uniquely assigns each record to a subject. You can
retrieve the registered schema back through the same call production uses — it is
nested under `head_primary_definition$schema`, exactly like the real API:

```{r}
data_source <- make_request(
  endpoint = paste0("api/warehouse/data-sources/", Sys.getenv("SOURCE_UUID"))
)
schema <- data_source$head_primary_definition$schema
```

### 2. Prepare and validate records

Records must comply with the schema. The sandbox validates them automatically at
ingestion, but you can validate up front with the same `jsonvalidate` /
`jsonlite` pattern the production pipeline uses. Be sure to set
`auto_unbox = TRUE` when serializing:

```{r}
records <- list(
  list(id = "xxxx", firstName = "John", lastName = "Doe",   age = 30, telephone = "1234567890"),
  list(id = "yyyy", firstName = "Jane", lastName = "Smith", age = 25, telephone = "0987654321")
)

json_schema <- jsonvalidate::json_schema$new(jsonlite::toJSON(schema, auto_unbox = TRUE))

validate_record <- function(record) {
  json_schema$validate(jsonlite::toJSON(record, auto_unbox = TRUE))
}

stopifnot(all(sapply(records, validate_record)))
```

### 3. Ingest records into the warehouse

Ingestion is a `POST` to the ingestion endpoint. The sandbox validates every
record against the registered schema before accepting it, stores accepted
payloads in memory, and writes a JSON copy to disk for
inspection. Supply the `source` (the same `Sys.getenv("SOURCE_UUID")` your
production app uses), the `records`, and optionally the `subject_field` that
links each record to an athlete.

```{r}
result <- make_request(
  endpoint = "api/warehouse/ingestion/primary/",
  method   = "POST",
  body = list(
    source        = Sys.getenv("SOURCE_UUID"),
    records       = records,
    subject_field = "id"
  )
)
```

Records that fail schema validation are **rejected** with an error naming exactly
which records failed and why — so you exercise the same failure handling you
would in production:

```{r}
# Missing the required `id` field -> rejected before anything is stored
bad <- list(list(firstName = "No", lastName = "Id"))
make_request(
  endpoint = "api/warehouse/ingestion/primary/",
  method   = "POST",
  body     = list(source = Sys.getenv("SOURCE_UUID"), records = bad, subject_field = "id")
)
#> Error: ... validation failed for record(s) 1 ...
```

### 4. Pull records back from the warehouse

Retrieval is a `GET` on the data-records endpoint. The only required query
parameter is `source_uuid`; set `paginate = TRUE` to return all matching records
(the sandbox returns a single page, mirroring the paginated shape).

```{r}
records <- make_request(
  endpoint = "api/warehouse/data-records",
  query    = list(source_uuid = Sys.getenv("SOURCE_UUID")),
  paginate = TRUE
)
```

Each returned record is wrapped in the same envelope the real API uses
(`id`, `dataset_uuid`, `data`, `subject`, `created_at`, `updated_at`). If a
record's `subject_field` value matches an athlete registered with
`create_profile()`, the sandbox resolves the `subject` to that athlete (id, name,
sport); otherwise `subject` is `NULL`. Resolution happens at **read time**, so
athletes registered *after* ingestion backfill on the next read.

## Inspecting and resetting the sandbox

Accepted payloads are written as pretty-printed JSON, one folder per data source,
so you can inspect the exact JSON that would have been sent to the API:

```{r}
browse_sandbox(Sys.getenv("SOURCE_UUID")) # opens the payload folder in the file explorer
browse_sandbox()                          # opens the sandbox root
```

To reset state between tests, clear a single source or the entire sandbox. This
also removes the on-disk payload files, and is convenient as test teardown:

```{r}
clear_sandbox(Sys.getenv("SOURCE_UUID")) # clear one source
clear_sandbox()                          # clear everything (schemas, records, orgs, profiles)
```

Sandbox state lasts only for the R session.

## A complete end-to-end example

Putting it all together — registration and warehouse workflows in one script,
all local, no network:

```{r}
library(csiapps)

# 1. (Optional) simulate login for a wrapped app's /me header
Sys.setenv(CSIAPPS_ACCESS_TOKEN = "your-dev-access-token")
set_institute("csiontario")
check_secrets()

# 2. Point the app at a placeholder source (CSIAPPS sets the real one in production)
Sys.setenv(SOURCE_UUID = "dev-placeholder")

# 3. Seed the registration registry
org <- create_sport_org("Rowing Canada")
create_profile(3, org$id, first_names = c("Ada", "Blair", "Cai"),
                          last_names  = c("Nkemelu", "Okafor", "Zhang"))

# 4. Read it back like production
orgs     <- fetch_org_options()
athletes <- fetch_profiles(filters = list(sport_org_id = org$id))

# 5. Register a warehouse schema and ingest records keyed to those athletes
register_sandbox_schema(Sys.getenv("SOURCE_UUID"), '{
  "type": "object",
  "required": ["athlete_id", "weight"],
  "properties": {
    "athlete_id": {"type": "integer"},
    "weight":     {"type": "number"}
  }
}')

make_request(
  endpoint = "api/warehouse/ingestion/primary/",
  method   = "POST",
  body = list(
    source        = Sys.getenv("SOURCE_UUID"),
    records       = list(
      list(athlete_id = athletes[[1]]$id, weight = 72.5),
      list(athlete_id = athletes[[2]]$id, weight = 68.1)
    ),
    subject_field = "athlete_id"
  )
)

# 6. Pull the records back; subjects resolve to the registered athletes
make_request(
  endpoint = "api/warehouse/data-records",
  query    = list(source_uuid = Sys.getenv("SOURCE_UUID")),
  paginate = TRUE
)

# 7. Reset when done
clear_sandbox()
```

When this code is deployed, disable sandbox mode and let the CSIAPPS team set
`SOURCE_UUID` to the real warehouse uuid they provision — the app code above is
**unchanged**. The developer never handles the real uuid, and the schema
registration in step 5 is development-only scaffolding: in production the schema
already lives in the remote warehouse.

## Worked example: integrating third-party vendor data

A common real task is joining a **third-party vendor's** data — a testing
platform, a wearable, a lab system — onto CSIAPPS athletes. CSI Ontario maintains
an **AMS_mapping** data source whose records link each CSIAPPS athlete (`id`) to
that athlete's profile in a vendor system (`vendor_profile_id`,
`vendor_profile_name`). A consumer app reads AMS_mapping to translate the
vendor's own ids into canonical CSIAPPS athlete ids.

In production the warehouse stores append-only mapping history and response
metadata. Consumer apps need only the current four-column mapping, so
`fetch_ams_mapping()` hides that difference: in the sandbox it returns the
simple current mapping you ingest; in production it flattens the full response,
keeps the latest record per mapping identity, and removes inactive mappings.

> Steps 1–4 below (dummy athletes, schema registration, CSV ingestion) are
> **development-only scaffolding**. In production the AMS_mapping source is
> already populated and maintained upstream; only the read-and-join in step 5 is
> real app code, and it ships **unchanged**.

### 1. Create dummy athletes and note their ids

The `id` column of your mapping must match real athlete ids. In the sandbox those
come from `create_profile()`, which assigns ids in insertion order.

A project often reads several named sources, so give each one its **own
descriptively-named** environment variable — here `AMS_MAPPING_UUID` — rather than
a generic `SOURCE_UUID`, so the app references each source unambiguously. As
before, CSIAPPS sets the real value in production; you use any placeholder in
development.

```{r}
Sys.setenv(AMS_MAPPING_UUID = "dev-ams-mapping") # CSIAPPS sets the real AMS_mapping uuid in production

org <- create_sport_org("Rowing Canada")
create_profile(3, org$id,
               first_names = c("Ada", "Blair", "Cai"),
               last_names  = c("Nkemelu", "Okafor", "Zhang"))

athletes <- fetch_profiles(filters = list(sport_org_id = org$id))
vapply(athletes, function(p) p$id, integer(1)) # e.g. 1 2 3 -- the ids the mapping must use
```

### 2. Author the AMS_mapping CSV

The CSV bridges two id systems: `id` is the CSIAPPS athlete id (matching the dummy
profiles above), and `vendor_profile_id` is the athlete's id in **your** vendor
system — a real value from your data (a code, integer, or UUID). The sandbox CSV
needs only the four columns that consumer apps use.

```{r}
mapping_csv <- tempfile(fileext = ".csv")
write.csv(
  data.frame(
    id                  = c(1, 2, 3),                        # CSIAPPS athlete ids (dummy profiles)
    vendor              = "VendorX",
    vendor_profile_id   = c("VX-8841", "VX-8842", "VX-9001"), # your real vendor ids
    vendor_profile_name = c("Ada N.", "Blair O.", "Cai Z."),
    stringsAsFactors    = FALSE
  ),
  mapping_csv, row.names = FALSE
)
```

### 3. Register the AMS_mapping schema

Register a development-only schema containing just the current mapping fields.
The sandbox does not need the production warehouse's `active` history or
response metadata:

```{r}
mapping_schema <- list(
  type = "object",
  additionalProperties = FALSE,
  required = as.list(c("id", "vendor", "vendor_profile_id", "vendor_profile_name")),
  properties = list(
    id                  = list(type = "number"),
    vendor              = list(type = "string"),
    vendor_profile_id   = list(type = "string"),
    vendor_profile_name = list(type = "string")
  )
)
register_sandbox_schema(Sys.getenv("AMS_MAPPING_UUID"), mapping_schema)
```

### 4. Ingest the CSV into the sandbox warehouse

Read the CSV back into records — forcing `vendor_profile_id` to stay a **string**
(vendor id formats vary) — then ingest with `subject_field = "id"` so each
mapping links to its CSIAPPS athlete:

```{r}
raw <- read.csv(mapping_csv,
                colClasses       = c(vendor_profile_id = "character"),
                stringsAsFactors = FALSE)

records <- lapply(seq_len(nrow(raw)), function(i) list(
  id                  = raw$id[i],
  vendor              = raw$vendor[i],
  vendor_profile_id   = raw$vendor_profile_id[i],
  vendor_profile_name = raw$vendor_profile_name[i]
))

make_request(
  endpoint = "api/warehouse/ingestion/primary/",
  method   = "POST",
  body = list(
    source        = Sys.getenv("AMS_MAPPING_UUID"),
    records       = records,
    subject_field = "id"
  )
)
```

### 5. Fetch the current mapping and join your vendor data

Set `AMS_MAPPING_UUID` to the UUID of the AMS mapping data source before
calling `fetch_ams_mapping()`. The function uses this environment variable
when `source_uuid` is not supplied and returns `id`, `vendor`,
`vendor_profile_id`, and `vendor_profile_name`. It stops with a clear error if
the UUID is missing or empty.

In production, `fetch_ams_mapping()` fetches the data-source records and
flattens and cleans the warehouse response so the result contains the most
up-to-date mapping. In sandbox mode, it returns the four columns you ingested
directly. Filter the result to your vendor, then `merge()` your real vendor
data onto the mapping by `vendor_profile_id`:

```{r}
mapping <- fetch_ams_mapping()
mapping <- mapping[mapping$vendor == "VendorX", ]

# your REAL third-party data, keyed by the vendor's own id
vendor_data <- data.frame(
  vendor_profile_id = c("VX-8841", "VX-8842", "VX-9001"),
  resting_hr        = c(52, 58, 49),
  stringsAsFactors  = FALSE
)

joined <- merge(vendor_data, mapping, by = "vendor_profile_id")
joined$athlete <- vapply(joined$id, function(i) {
  p <- fetch_profile(profile_id = i)
  paste(p$person$first_name, p$person$last_name)
}, character(1))
joined # vendor_profile_id | resting_hr | id | vendor | vendor_profile_name | athlete
```

### Going to production

Disable sandbox mode and ensure `AMS_MAPPING_UUID` contains the real AMS
mapping data-source UUID. The same `fetch_ams_mapping()` call will then return
the cleaned, current production mapping.

## Limitations

The sandbox faithfully simulates the *schema contract*, not the warehouse.
Anything that depends on server-side state will differ from production:

-   **Validation parity is approximate.** The sandbox validates records against
    the JSON Schema only (with Ajv, via `jsonvalidate`), catching the most common
    failures: missing required fields, wrong types, length and pattern
    violations. The real server additionally enforces things the schema cannot
    express — `subject_field` resolution against registered profiles, duplicate
    and dataset handling against existing data, and token permissions — and
    validators may differ on edge cases (e.g. `format` enforcement). Passing
    sandbox validation is therefore a *necessary but not sufficient* condition
    for production acceptance; do not treat a green sandbox run as a guarantee.

-   **`subject` linkage is emulated only for registered athletes.** On retrieval
    the sandbox resolves each record's `subject_field` value against athletes
    created with `create_profile()`, returning the matched athlete — or `NULL` if
    none is registered. Production would reject an unresolved subject; the
    sandbox accepts it silently.

-   **Only warehouse endpoints go through `make_request()`.** Registration
    endpoints raise a 501; use the `fetch_*` helpers, which read the dummy
    registry. Any other endpoint is unsupported.

See `?"csiapps-sandbox"` for the full reference.
